Return on investment of private sector malaria control at a large sugar facility in Southern Mozambique

### 2018-04-11 inst/rmd/00_title_page.Rmd
major changes

This paper provides new empirical evidence regarding the return on investment of privately managed malaria control activities (indoor residual spraying with pesticides) on worker absenteeism in Mozambique. We analyze 4 years of malaria control and worker health and absenteeism data from a large sugar processing facility in Mozambique. We find that the benefits outweight the costs (ie, there is a positive return on investment) even when the consideration of benefits is limited to those directly accrued by the company. These findings suggest that the private sector may have an important role to play in malaria control in endemic areas.

>>>>

This paper provides new empirical evidence regarding the effect and return on investment of privately managed malaria control activities (indoor residual spraying with pesticides) on worker absenteeism in Mozambique. We analyze 4 years of malaria control and worker health and absenteeism data from a large sugar processing facility in Mozambique. We find that the benefits outweight the costs (ie, there is a positive return on investment) even when the consideration of benefits is limited to those directly accrued by the company. These findings suggest that the private sector may have an important role to play in malaria control in endemic areas.

### 2018-04-11 inst/rmd/01_introduction.Rmd
formatting updates
major changes

Malaria has a large economic impact on endemic societies. By affecting saving, investment [@Shretta2016], risk perception, productivity, absenteeism [@Nonvignon_2016], human capital accumulation [@CastelBranco2014], mortality, and costs of care [@Sachs2002], malaria likely has a negative effect on GDP and growth [@McCarthy_2000] [@Orem_2012]. Because of the relative affordability of most intereventions and the enormous societal costs of malaria, most forms of malaria control are cost-effective when a public welfare perspective is assumed, such as when a government provides the financing [@White_2011] [@Purdy_2013] [@Howard_2017].

@@

A significant sector of the economy in Mozambique is dominated by a full large-scale foreing direct investment projects [@Robbins2012], and the role of the private sector in health generally, and malaria specifically, is unequivocally important. Large agriculture and extractive industry firms take up wide swaths of land and employ hundreds of thousands  [@German2013]. The Mozambican state has encouraged large-scale entreprise with the aim of general economic development [@Buur2012]. And where large firms exist, they often take on social roles such as housing and health care [@Winkler]. At times, this role is necessary from a purely practical standpoint; in other cases, it is employed under the guise of "corporate social responsibility" [@Azemar2009]. Regardless of the language used, it is clear that private industry plays an important role in public health in Mozambique [@Robbins2012] [@CastelBranco2014].

Several cases exist of foreign firms engaging in large-scale malaria control campaigns [@Mouzin2011] [@Han] [@Bennett_2017] [@Kaula_2017]. But these studies generally consider population health as the outcome measure of interest, rather than worker absenteeism or productivity. Similarly, they often neglect to differentiate between those clinical costs which are absorbed by the local health system versus those which are absorbed by the firm itself. In the literature, making the "investment case" for malaria control or elimination generally implies that the investor is the public sector, and takes into account those costs and benefits which are applicable from a public welfare point of view [@Shretta_2017]; though appropriate in most cases (the government or institutions interested in public welfare primarily being the primary malaria control agents in most locations), the findings of these studies are rarely applicable to the private sector. In the case of a private firm not interested in "corporate social responsibility", it is not clear whether investing in malaria control would be profitable or not. This lack of clarity not only discourages investment, but also makes it difficult for governments to pinpoint the correct of amount of subsidy (if applicable) to encourage private sector scale-up in malaria control.

To address the question of the profitability of malaria control activities from the standpoint of a private firm, we analyze data during a 7 year period from a private sugar facility in Southern Mozambique. We assess the effect of indoor residual spraying (IRS) on the absenteeism and health of workers, and demonstrate that the firm's engagement in malaria control not only improved worker health, but also generated a positive return on investment.

The structure of the paper is as follows. After a brief review of the existing literature on the subject of private sector investment in malaria control and its effects on worker productivity, we provide an overview of the sugar company under study, and the epidemiology of malaria in the nearby area, as well as in Mozambique as a whole. We then give an overview of the data collected, and outline the theoretical and methdological assumptions that underly our analysis. In the results, we show the effectiveness of the company's malaria control program in reducing absences, and translate this reduction into cost savings. Our robustness checks consist of \todo{Add some stuff here}. Our discussion covers potential implications from this study in terms of policy and investment, as well as the paper's limitations.

@@

Details on Maragra:

Maragra Açucar SA (henceforth referred to as "Maragra")\todo{This is a comment that will appear in the margin}

>>>>

Malaria accounts for a half million annual deaths worldwide [@White2014] [@World2016] [@Ashley2018]. Though rapid improvements in technology and funding have lead to important reductions in marlaria's global burden, the scale-up in activities required for eradication ("the worldwide interruption of transmission") [@TheLancet2011] will mean new partnerships and actors. One promising - albeit atypical - potential stakeholder in global malaria eradication is the private sector, given its omnipresence and potential to benefit directly from the elimination of malaria. But little evidence exists demonstrating how private sector entities can engage with malaria control and benefit at the firm level.

At the societal level, malaria has a large economic impact. By affecting saving, investment [@Shretta2016], risk perception, productivity, absenteeism [@Nonvignon_2016], human capital accumulation [@CastelBranco2014], mortality, and costs of care [@Sachs2002], malaria likely has a negative effect on GDP and growth [@McCarthy_2000] [@Orem_2012]. Because of the relative affordability of most intereventions and the enormous societal costs of malaria, most forms of malaria control are cost-effective when a public welfare perspective is assumed, such as when a government provides the financing [@White_2011] [@Purdy_2013] [@Howard_2017].

@@

A significant sector of the economy in Mozambique is dominated by large-scale foreign direct investment projects [@Robbins2012], and the role of the private sector in health generally, and malaria specifically, is unequivocally important. Large agriculture and extractive industry firms take up wide swaths of land and employ hundreds of thousands  [@German2013]. The Mozambican state has encouraged large-scale entreprise with the aim of general economic development [@Buur2012]. And where large firms exist, they often take on social roles such as housing and health care [@Winkler]. At times, this role is necessary from a purely practical standpoint; in other cases, it is employed under the auspices of "corporate social responsibility" [@Azemar2009]. Regardless of the language used, it is clear that private industry plays an important role in public health in Mozambique [@Robbins2012] [@CastelBranco2014].

Several cases exist of foreign firms engaging in large-scale malaria control campaigns [@Mouzin2011] [@Han] [@Bennett_2017] [@Kaula_2017]. But these studies generally consider population health as the outcome measure of interest, rather than worker absenteeism or productivity. Similarly, they often neglect to differentiate between those clinical costs which are absorbed by the local health system versus those which are absorbed by the firm itself. In the literature, making the "investment case" for malaria control or elimination generally implies that the investor is the public sector, and takes into account those costs and benefits which are applicable from a public welfare point of view [@Shretta_2017]; though appropriate in most cases (the government or institutions interested in public welfare primarily being the primary malaria control agents in most locations), the findings of these studies are rarely applicable to the private sector, and even less so at granular levels (such as an individual firm). In the case of a private firm not interested in "corporate social responsibility", it is not clear whether investing in malaria control would be profitable or not. This lack of clarity not only discourages investment, but also makes it difficult for governments to pinpoint the correct of amount of subsidy (if applicable) to encourage private sector scale-up in malaria control.

To address the question of the profitability of malaria control activities from the standpoint of a private firm, we analyze data during a 4 year period from a private sugar facility in Southern Mozambique. We assess the effect of indoor residual spraying (IRS) on the absenteeism and health of workers, and demonstrate that the firm's engagement in malaria control not only improved worker health, but also generated a positive return on investment from a pure accounting perspective.

The structure of the paper is as follows. After a brief review of the existing literature on the subject of private sector investment in malaria control and its effects on worker productivity, we provide an overview of the sugar company under study, and the epidemiology of malaria in the nearby area, as well as in Mozambique as a whole. We then give an overview of the data collected, and outline the theoretical and methdological assumptions and tools that underly our analysis. In the results, we show the effectiveness of the company's malaria control program in reducing absences, and translate this reduction into cost savings. Our discussion covers potential implications from this study in terms of policy and investment, as well as the paper's limitations.

@@

Sugar has been systematically cultivated in Mozambique since the late 1800s, but was not exported at a large scale until the 1950s. The Incomati Estates company, a small sugarcane processing facility started by a Scotsman on the banks of the Incomati River in 1913, was the first firm to export sugar from Mozambique. Following its purchase by international investors in the 1950s, it (along with the rest of the industry) expanded significantly. In the late 1960s, a Portuguese family opened the Maragra Açúcar company, while a group of foreign investors started the nearby Marracuene Agrícola Açucareira mill. By the early 1970s, sugar grew to account for greater than 10% of Mozambique's national exports. Nationalized following independence in 1977, the industry's production levels fell from 320,000 annual tons to fewer than 15,000 in 1992. After the end of the civil war, foreign investment revived the sugar industry, and by 2011 production had surpassed its 1972 peak.

The mill of the Maragra Açucar SA (a subsidary of the Illovo sugar company, henceforth referred to as "Maragra") was nationalized in the 1970s (like all other mills), went through a period of low production, and then fell completely out of use by 1984. It re-opened in private hands in 1992, and was renovated by a group of international investors in 1998. Today, Maragra accounts for roughly one quarter of Mozambique's overall sugar production (second only to the nearby Xinavane mill run by the Hulett Sugar Tongaat company) [@sutton]. With a favorably close location to the port of Maputo, ample land, approximately 5,000 employees (of which three quarters are seasonal), and a mill with the capacity to process not only all the sugarcane grown on Maragra's land, but also the cane of the many nearby smallholders [@OLaughlin2016], Maragra has so far been able to weather the 2016 Mozambican crisis and concurrent collapse in global sugar prices.

### 2018-04-11 inst/rmd/paper.tex
major changes

Malaria has a large economic impact on endemic societies. By affecting
saving, investment (Shretta et al., 2016), risk perception,
productivity, absenteeism (Nonvignon et al., 2016), human capital
accumulation (Castel-Branco, 2014), mortality, and costs of care (Sachs
and Malaney, 2002), malaria likely has a negative effect on GDP and
growth (McCarthy et al., 2000) (Orem et al., 2012). Because of the
relative affordability of most intereventions and the enormous societal
costs of malaria, most forms of malaria control are cost-effective when
a public welfare perspective is assumed, such as when a government
provides the financing (White et al., 2011) (Purdy et al., 2013) (Howard
et al., 2017).

@@

A significant sector of the economy in Mozambique is dominated by a full
large-scale foreing direct investment projects (Robbins and Perkins,
2012), and the role of the private sector in health generally, and
malaria specifically, is unequivocally important. Large agriculture and
extractive industry firms take up wide swaths of land and employ
hundreds of thousands (German et al., 2013). The Mozambican state has
encouraged large-scale entreprise with the aim of general economic
development (Buur et al., 2012). And where large firms exist, they often
take on social roles such as housing and health care (Winkler, 2013). At
times, this role is necessary from a purely practical standpoint; in
other cases, it is employed under the guise of ``corporate social
responsibility'' (Azemar and Desbordes, 2009). Regardless of the
language used, it is clear that private industry plays an important role
in public health in Mozambique (Robbins and Perkins, 2012)
(Castel-Branco, 2014).

Several cases exist of foreign firms engaging in large-scale malaria
control campaigns (Mouzin and al., 2011) (Han, 2015) (Bennett et al.,
2017) (Kaula et al., 2017). But these studies generally consider
population health as the outcome measure of interest, rather than worker
absenteeism or productivity. Similarly, they often neglect to
differentiate between those clinical costs which are absorbed by the
local health system versus those which are absorbed by the firm itself.
In the literature, making the ``investment case'' for malaria control or
elimination generally implies that the investor is the public sector,
and takes into account those costs and benefits which are applicable
from a public welfare point of view (Shretta et al., 2017); though
appropriate in most cases (the government or institutions interested in
public welfare primarily being the primary malaria control agents in
most locations), the findings of these studies are rarely applicable to
the private sector. In the case of a private firm not interested in
``corporate social responsibility'', it is not clear whether investing
in malaria control would be profitable or not. This lack of clarity not
only discourages investment, but also makes it difficult for governments
to pinpoint the correct of amount of subsidy (if applicable) to
encourage private sector scale-up in malaria control.

To address the question of the profitability of malaria control
activities from the standpoint of a private firm, we analyze data during
a 7 year period from a private sugar facility in Southern Mozambique. We
assess the effect of indoor residual spraying (IRS) on the absenteeism
and health of workers, and demonstrate that the firm's engagement in
malaria control not only improved worker health, but also generated a
positive return on investment.

The structure of the paper is as follows. After a brief review of the
existing literature on the subject of private sector investment in
malaria control and its effects on worker productivity, we provide an
overview of the sugar company under study, and the epidemiology of
malaria in the nearby area, as well as in Mozambique as a whole. We then
give an overview of the data collected, and outline the theoretical and
methdological assumptions that underly our analysis. In the results, we
show the effectiveness of the company's malaria control program in
reducing absences, and translate this reduction into cost savings. Our
robustness checks consist of \todo{Add some stuff here}. Our discussion
covers potential implications from this study in terms of policy and
investment, as well as the paper's limitations.

@@

Details on Maragra:

Maragra Açucar SA (henceforth referred to as ``Maragra'').

@@

\hypertarget{refs}{}
\hypertarget{ref-Azemar2009}{}
Azemar, C., Desbordes, R., 2009. Public governance, health and foreign
direct investment in sub-saharan africa. Journal of African Economies
18, 667--709. \url{https://doi.org/10.1093/jae/ejn028}

@@

@@

@@

@@

>>>>

Malaria accounts for a half million annual deaths worldwide (White et
al., 2014) (WHO, 2016) (Ashley et al., 2018). Though rapid improvements
in technology and funding have lead to important reductions in
marlaria's global burden, the scale-up in activities required for
eradication (``the worldwide interruption of transmission'') (Lancet,
2011) will mean new partnerships and actors. One promising - albeit
atypical - potential stakeholder in global malaria eradication is the
private sector, given its omnipresence and potential to benefit directly
from the elimination of malaria. But little evidence exists
demonstrating how private sector entities can engage with malaria
control and benefit at the firm level.

At the societal level, malaria has a large economic impact. By affecting
saving, investment (Shretta et al., 2016), risk perception,
productivity, absenteeism (Nonvignon et al., 2016), human capital
accumulation (Castel-Branco, 2014), mortality, and costs of care (Sachs
and Malaney, 2002), malaria likely has a negative effect on GDP and
growth (McCarthy et al., 2000) (Orem et al., 2012). Because of the
relative affordability of most intereventions and the enormous societal
costs of malaria, most forms of malaria control are cost-effective when
a public welfare perspective is assumed, such as when a government
provides the financing (White et al., 2011) (Purdy et al., 2013) (Howard
et al., 2017).

@@

A significant sector of the economy in Mozambique is dominated by
large-scale foreign direct investment projects (Robbins and Perkins,
2012), and the role of the private sector in health generally, and
malaria specifically, is unequivocally important. Large agriculture and
extractive industry firms take up wide swaths of land and employ
hundreds of thousands (German et al., 2013). The Mozambican state has
encouraged large-scale entreprise with the aim of general economic
development (Buur et al., 2012). And where large firms exist, they often
take on social roles such as housing and health care (Winkler, 2013). At
times, this role is necessary from a purely practical standpoint; in
other cases, it is employed under the auspices of ``corporate social
responsibility'' (Azemar and Desbordes, 2009). Regardless of the
language used, it is clear that private industry plays an important role
in public health in Mozambique (Robbins and Perkins, 2012)
(Castel-Branco, 2014).

Several cases exist of foreign firms engaging in large-scale malaria
control campaigns (Mouzin and al., 2011) (Han, 2015) (Bennett et al.,
2017) (Kaula et al., 2017). But these studies generally consider
population health as the outcome measure of interest, rather than worker
absenteeism or productivity. Similarly, they often neglect to
differentiate between those clinical costs which are absorbed by the
local health system versus those which are absorbed by the firm itself.
In the literature, making the ``investment case'' for malaria control or
elimination generally implies that the investor is the public sector,
and takes into account those costs and benefits which are applicable
from a public welfare point of view (Shretta et al., 2017); though
appropriate in most cases (the government or institutions interested in
public welfare primarily being the primary malaria control agents in
most locations), the findings of these studies are rarely applicable to
the private sector, and even less so at granular levels (such as an
individual firm). In the case of a private firm not interested in
``corporate social responsibility'', it is not clear whether investing
in malaria control would be profitable or not. This lack of clarity not
only discourages investment, but also makes it difficult for governments
to pinpoint the correct of amount of subsidy (if applicable) to
encourage private sector scale-up in malaria control.

To address the question of the profitability of malaria control
activities from the standpoint of a private firm, we analyze data during
a 4 year period from a private sugar facility in Southern Mozambique. We
assess the effect of indoor residual spraying (IRS) on the absenteeism
and health of workers, and demonstrate that the firm's engagement in
malaria control not only improved worker health, but also generated a
positive return on investment from a pure accounting perspective.

The structure of the paper is as follows. After a brief review of the
existing literature on the subject of private sector investment in
malaria control and its effects on worker productivity, we provide an
overview of the sugar company under study, and the epidemiology of
malaria in the nearby area, as well as in Mozambique as a whole. We then
give an overview of the data collected, and outline the theoretical and
methdological assumptions and tools that underly our analysis. In the
results, we show the effectiveness of the company's malaria control
program in reducing absences, and translate this reduction into cost
savings. Our discussion covers potential implications from this study in
terms of policy and investment, as well as the paper's limitations.

@@

Sugar has been systematically cultivated in Mozambique since the late
1800s, but was not exported at a large scale until the 1950s. The
Incomati Estates company, a small sugarcane processing facility started
by a Scotsman on the banks of the Incomati River in 1913, was the first
firm to export sugar from Mozambique. Following its purchase by
international investors in the 1950s, it (along with the rest of the
industry) expanded significantly. In the late 1960s, a Portuguese family
opened the Maragra Açúcar company, while a group of foreign investors
started the nearby Marracuene Agrícola Açucareira mill. By the early
1970s, sugar grew to account for greater than 10\% of Mozambique's
national exports. Nationalized following independence in 1977, the
industry's production levels fell from 320,000 annual tons to fewer than
15,000 in 1992. After the end of the civil war, foreign investment
revived the sugar industry, and by 2011 production had surpassed its
1972 peak.

The mill of the Maragra Açucar SA (a subsidary of the Illovo sugar
company, henceforth referred to as ``Maragra'') was nationalized in the
1970s (like all other mills), went through a period of low production,
and then fell completely out of use by 1984. It re-opened in private
hands in 1992, and was renovated by a group of international investors
in 1998. Today, Maragra accounts for roughly one quarter of Mozambique's
overall sugar production (second only to the nearby Xinavane mill run by
the Hulett Sugar Tongaat company) (Sutton, 2015). With a favorably close
location to the port of Maputo, ample land, approximately 5,000
employees (of which three quarters are seasonal), and a mill with the
capacity to process not only all the sugarcane grown on Maragra's land,
but also the cane of nearby smallholders, Maragra has so far been able
to weather the 2016 Mozambican crisis and concurrent collapse in global
sugar prices.

\begin{itemize}
\tightlist
\item
  Details on Maragra: .
\end{itemize}

-Xinavane (O'Laughlin, 2016)

@@

\hypertarget{refs}{}
\hypertarget{ref-Ashley2018}{}
Ashley, E.A., Phyo, A.P., Woodrow, C.J., 2018. Malaria. The Lancet.
\url{https://doi.org/10.1016/s0140-6736(18)30324-6}

\hypertarget{ref-Azemar2009}{}
Azemar, C., Desbordes, R., 2009. Public governance, health and foreign
direct investment in sub-saharan africa. Journal of African Economies
18, 667--709. \url{https://doi.org/10.1093/jae/ejn028}

@@

\hypertarget{ref-TheLancet2011}{}
Lancet, T., 2011. Malaria: Control vs elimination vs eradication. The
Lancet 378, 1117. \url{https://doi.org/10.1016/s0140-6736(11)61489-x}

@@

\hypertarget{ref-OLaughlin2016}{}
O'Laughlin, B., 2016. Consuming bodies: Health and work in the cane
fields in xinavane, mozambique. Journal of Southern African Studies 43,
625--641. \url{https://doi.org/10.1080/03057070.2016.1190519}

@@

\hypertarget{ref-sutton}{}
Sutton, J., 2015. Mapa empresarial de moçambique.

@@

\hypertarget{ref-White2014}{}
White, N.J., Pukrittayakamee, S., Hien, T.T., Faiz, M.A., Mokuolu, O.A.,
Dondorp, A.M., 2014. Malaria. The Lancet 383, 723--735.
\url{https://doi.org/10.1016/s0140-6736(13)60024-0}

\hypertarget{ref-World2016}{}
WHO, 2016. World Malaria Report.
